\begin{thebibliography}{9}

\bibitem{lake1989enhanced}
Lake, L. W. (1989). \emph{Enhanced oil recovery}. Prentice Hall.

\bibitem{bear1972dynamics}
Bear, J. (1972). \emph{Dynamics of fluids in porous media}. American Elsevier Publishing Company.

\bibitem{saffman1958penetration}
Saffman, P. G., \& Taylor, G. I. (1958). The penetration of a fluid into a porous medium or Hele-Shaw cell containing a more viscous liquid. \emph{Proceedings of the Royal Society of London. Series A. Mathematical and Physical Sciences}, 245(1242), 312-329.

\bibitem{homsy1987viscous}
Homsy, G. M. (1987). Viscous fingering in porous media. \emph{Annual Review of Fluid Mechanics}, 19(1), 271-311.

\bibitem{sheng2011modern}
Sheng, J. J. (2011). \emph{Modern chemical enhanced oil recovery: theory and practice}. Gulf Professional Publishing.

\bibitem{binks2002particles}
Binks, B. P. (2002). Particles as surfactants—similarities and differences. \emph{Current Opinion in Colloid \& Interface Science}, 7(1-2), 21-41.

\bibitem{jackson1993hydrodynamics}
Jackson, D. P., Goldstein, R. E., \& Cebers, A. (1993). Hydrodynamics of fingering instabilities in dipolar fluids. \emph{Physical Review E}, 47(6), 3981.

\bibitem{miranda1998radial}
Miranda, J. A., \& Widom, M. (1998). Radial viscous fingering of a magnetic fluid in a weakly applied magnetic field. \emph{Physical Review E}, 58(2), 2131.

\bibitem{sorbie1991polymer}
Sorbie, K. S. (1991). \emph{Polymer-improved oil recovery}. Springer Science \& Business Media.

\bibitem{olajire2014review}
Olajire, A. A. (2014). Review of ASP EOR (alkaline surfactant polymer enhanced oil recovery) technology in the petroleum industry: Prospects and challenges. \emph{Energy}, 77, 963-982.

% --- Mergeado de LBM-Teoria.tex ---

\bibitem{chen1998lattice}
Chen, S., \& Doolen, G. D. (1998). Lattice Boltzmann method for fluid flows. \emph{Annual Review of Fluid Mechanics}, 30(1), 329-364.

\bibitem{qian1992lattice}
Qian, Y. H., d'Humières, D., \& Lallemand, P. (1992). Lattice BGK models for Navier-Stokes equation. \emph{EPL (Europhysics Letters)}, 17(6), 479.

\bibitem{bhatnagar1954model}
Bhatnagar, P. L., Gross, E. P., \& Krook, M. (1954). A model for collision processes in gases. I. Small amplitude processes in charged and neutral one-component systems. \emph{Physical Review}, 94(3), 511.

\bibitem{guo2002discrete}
Guo, Z., Zheng, C., \& Shi, B. (2002). Discrete lattice effects on the forcing term in the lattice Boltzmann method. \emph{Physical Review E}, 65(4), 046308.

\bibitem{cahn1958free}
Cahn, J. W., \& Hilliard, J. E. (1958). Free energy of a nonuniform system. I. Interfacial free energy. \emph{The Journal of Chemical Physics}, 28(2), 258-267.

\bibitem{zou1997analytical}
Zou, Q., \& He, X. (1997). On pressure and velocity boundary conditions for the lattice Boltzmann BGK model. \emph{Physics of Fluids}, 9(6), 1591-1598.

\bibitem{brinkman1949calculation}
Brinkman, H. C. (1949). A calculation of the viscous force exerted by a flowing fluid on a dense swarm of particles. \emph{Flow, Turbulence and Combustion}, 1(1), 27-34.

\bibitem{ladd1994numerical}
Ladd, A. J. C. (1994). Numerical simulations of particulate suspensions via a discretized Boltzmann equation. Part 1. Theoretical foundation. \emph{Journal of Fluid Mechanics}, 271, 285-309.

\bibitem{ziegler1993boundary}
Ziegler, D. P. (1993). Boundary conditions for lattice Boltzmann simulations. \emph{Journal of Statistical Physics}, 71(5), 1171-1177.

\bibitem{white2006viscous}
White, F. M. (2006). \emph{Viscous Fluid Flow} (3rd ed.). McGraw-Hill.

\bibitem{gontijo_cunha_review}
Gontijo, R. G., \& Cunha, F. R. (2015). Numerical simulations of magnetic suspensions with hydrodynamic and dipole-dipole magnetic interactions. \emph{Physics of Fluids}, 27(12), 122002.

\bibitem{rosensweig1985ferrohydrodynamics}
Rosensweig, R. E. (1985). \emph{Ferrohydrodynamics}. Cambridge University Press.

% --- Citadas mas ainda sem entrada (preencher) ---
% whitaker1999method, batchelor1967introduction, stokes1851effect,
% dullien1992porous


\end{thebibliography}
